Supplement: Autopolyploid establishment through polygenic adaptation

### 2025-01-14 doc/paper/supp.tex
update
paper update

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[m-]{paper}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thealgorithm}{S\arabic{algorithm}}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
\caption{
Validation of the autotetraploid and triploid infinitesimal model by a
comparison against the discrete locus model with $L=500$ additive loci, showing
the average phenotypic variance in each generation averaged over 10 replicate
simulations for both models.
We assume the initial phenotypic variance to be one for all simulations, and
all replicates are initialized randomly in accordance with this initial
phenotypic variance.
Allelic effects for the discrete locus model are sampled from a Gaussian with
mean $0$ and variance $1/2L$.
\label{fig:fininf}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
\caption{Comparison of the mixed-ploidy infinitesimal model with the
\(L\)-locus model, for \(L=500,100,50\) and \(20\). The decline in the
genetic variance \(V_z\) within each cytotype due to drift is shown. The
transparent lines show the complete simulation, whereas the solid line
shows the same data but smoothed in overlapping windows of 20
generations. We assume \(N=500, u=v=0.08\) and no selection. In the top
row where \(\beta_k^2=1, \alpha_k=0\), the equilibrium variance in the
absence of inbreeding in triploids is 2/3 that of diploids, and in
tetraploids it is twice that in diploids. In the middle row,
\(\beta_3^2=2/3\) and \(\beta_4^2=1/2\), so that the equilibrium
variance in the absence of inbreeding is equal across cytotypes. In the
bottom row, \(\alpha_2= \alpha_3 =1/2\) and \(\alpha_4=1/6\), causing an
immediate increase in the genetic variance in higher cytotypes, but also
accelerated inbreeding. \label{fig:vz}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
\caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
a mixed-ploidy population for different values of \(\alpha\) (we assume
\(\alpha_k=\alpha\), where \(\alpha_k\) is the probability that a
diploid gamete produced by a \(k\)-ploid cytotype contains two copies of
the same parental gene at a random locus). We assume \(N=500, u=v=0.08\)
and no selection.}\label{fig:f1}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/g1diff.pdf}
\caption{
Deterministic mixed-ploidy equilibrium. The left plot shows the stable (solid
lines) and unstable (dashed lines) equilibria for the proportion of haploid
gametes in the gamete pool $g_1$ as a function of $u$ for different values of
$v$. The middle plot shows the relationship between $\Delta g_1 = g_1' - g_1$
and $g_1$. The zeros of this graph are the fixed points of the dynamical system
and are indicated by the hollow (unstable equilibrium) and solid (stable
equilibrium) dots. The rightmost plot shows the region of parameter space where
a stable mixed-ploidy equilibrium exist.
\label{fig:g1diff}}
\end{figure}

@@

For finite $N$, the basic mixed-ploidy model defines a Markov chain on the
state space $[0..N] \times [0..N]$.
\begin{align}
  p_{ij,kl} = \Pr\{N_2(t+1)=k,& N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
    &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
\end{align}
where 
\begin{align}
  p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
  p_3 &= \left(\frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
    {N(1-u) + (i+j)u + j(2v-1)}\right)^2 
\end{align}
Associating a unique index with each pair $(i,j)$ with $0 \le i,j \le N$, we
can define a transition probability matrix $P$ of dimensions $(N+1)^2 \times
(N+1)^2$ for this Markov chain.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
\caption{Expected time to diploid ancestry. The solid lines show
\(\Ex[T_4]\), i.e.~the expected time since being inherited from a
diploid ancestor for a random gene in a tetraploid individual at
equilibrium, for different values of \(v\) (half the triploid
fertility). The dashed lines show \(\Ex[T_3]\), i.e.~the same quantity
for a gene sampled from a triploid. Note that \(\Ex[T]\) blows up
whenever \(u\) and \(v\) exceed their critical value for tetraploid
establishment. \label{fig:ttdip}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
\caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
in the diploid subpopulation are shown for different values of \(u\) and
associated values of \(N\), keeping \(N_e = (1-2u)N\) constant at 200.
We assume \(u=v\). All lines coincide almost completely and are
indistinguishable from \(1-e^{-t/2N_e}\). Results are shown for
\(\alpha_k=1/6\). As \(\alpha_k\) decreases to 0, \(\bar{F}\) in the
mixed-ploidy population becomes completely indistinguishable from
\(\bar{F}\) in the diploid subpopulation.}\label{fig:fNe}
\end{figure}

@@

Experimental results indicate that, at least in yeast, triploids usually form
trivalents and undergo recombination, after which each trivalent is randomly
assorted in the daughter cells, some receiving one, others two copies of a
given chromosome \citep{charles2010}.
In the absence of gametic nonreduction, the probability
of obtaining euploid gametes (two diploid and two haploid gametes) from such 
a process is $(1/2)^n$, where $n$ is the number of chromosomes. If the number
of chromosomes is small this is not negligible, for instance in
\textit{Arabidopsis thaliana} we would have $(1/2)^5 \approx 0.03$, which is of
the same order as the unreduced gamete formation rate.
Unreduced (triploid) gametes may also be produced and important for the
dynamics of mixed-ploidy populations \citep{ramsey1998}.
However, they generate additional difficulty, since in order to compute the
contributed variance under inbreeding, we would need an additional identity
coefficient recording the probability that three genes are IBD at a locus.
We will hence ignore the possibility of unreduced gamete production in
triploids.
We note that, on the supposition that diploid gametes are produced by random
assortment of chromosomes in a haploid and diploid gametes, $\alpha_3 \le 1/4$.

@@

\bibliographystyle{abbrvnat}
\bibliography{/home/arthur_z/vimwiki/bib.bib}
\end{document}

>>>>

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[m-]{paper}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
\caption{
Validation of the autotetraploid and triploid infinitesimal model by a
comparison against the discrete locus model with $L=500$ additive loci, showing
the average phenotypic variance in each generation averaged over 10 replicate
simulations for both models.
We assume the initial phenotypic variance to be one for all simulations, and
all replicates are initialized randomly in accordance with this initial
phenotypic variance.
Allelic effects for the discrete locus model are sampled from a Gaussian with
mean $0$ and variance $1/2L$.
\label{fig:fininf}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
\caption{Comparison of the mixed-ploidy infinitesimal model with the
\(L\)-locus model, for \(L=500,100,50\) and \(20\). The decline in the
genetic variance \(V_z\) within each cytotype due to drift is shown. The
transparent lines show the complete simulation, whereas the solid line
shows the same data but smoothed in overlapping windows of 20
generations. We assume \(N=500, u=v=0.08\) and no selection. In the top
row where \(\beta_k^2=1, \alpha_k=0\), the equilibrium variance in the
absence of inbreeding in triploids is 2/3 that of diploids, and in
tetraploids it is twice that in diploids. In the middle row,
\(\beta_3^2=2/3\) and \(\beta_4^2=1/2\), so that the equilibrium
variance in the absence of inbreeding is equal across cytotypes. In the
bottom row, \(\alpha_2= \alpha_3 =1/2\) and \(\alpha_4=1/6\), causing an
immediate increase in the genetic variance in higher cytotypes, but also
accelerated inbreeding. \label{fig:vz}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
\caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
a mixed-ploidy population for different values of \(\alpha\) (we assume
\(\alpha_k=\alpha\), where \(\alpha_k\) is the probability that a
diploid gamete produced by a \(k\)-ploid cytotype contains two copies of
the same parental gene at a random locus). We assume \(N=500, u=v=0.08\)
and no selection.}\label{fig:f1}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr-check.pdf}
\caption{
    Nonzero $\alpha_k$ parameters increase the tetraploid establishment
    probability, but mainly due to $\alpha_2$, which increases the segregation
    variance associated with the formatoin of diploid gametes by diploids.
    All other parameters and simulation details are as in \cref{m-fig:estwmig}.
\label{fig:suppdr}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-13/g1diff.pdf}
\caption{
Deterministic mixed-ploidy equilibrium. The left plot shows the stable (solid
lines) and unstable (dashed lines) equilibria for the proportion of haploid
gametes in the gamete pool $g_1$ as a function of $u$ for different values of
$v$. The middle plot shows the relationship between $\Delta g_1 = g_1' - g_1$
and $g_1$. The zeros of this graph are the fixed points of the dynamical system
and are indicated by the hollow (unstable equilibrium) and solid (stable
equilibrium) dots. The rightmost plot shows the region of parameter space where
a stable mixed-ploidy equilibrium exist.
\label{fig:g1diff}}
\end{figure}

@@

For finite $N$, the basic mixed-ploidy model defines a Markov chain on the
state space $[0..N] \times [0..N]$.
\begin{align}
  p_{ij,kl} &= \Pr\{N_2(t+1)=k, N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
    &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
\end{align}
where 
\begin{align}
  p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
  p_3 &= \left(\frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
    {N(1-u) + (i+j)u + j(2v-1)}\right)^2 
\end{align}
Associating a unique index with each pair $(i,j)$ with $0 \le i,j \le N$, we
can define a transition probability matrix $P$ of dimensions $(N+1)^2 \times
(N+1)^2$ for this Markov chain.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
\caption{Expected time to diploid ancestry. The solid lines show
\(\Ex[T_4]\), i.e.~the expected time since being inherited from a
diploid ancestor for a random gene in a tetraploid individual at
equilibrium, for different values of \(v\) (half the triploid
fertility). The dashed lines show \(\Ex[T_3]\), i.e.~the same quantity
for a gene sampled from a triploid. Note that \(\Ex[T]\) blows up
whenever \(u\) and \(v\) exceed their critical value for tetraploid
establishment. \label{fig:ttdip}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
\caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
in the diploid subpopulation are shown for different values of \(u\) and
associated values of \(N\), keeping \(N_e = (1-2u)N\) constant at 200.
We assume \(u=v\). All lines coincide almost completely and are
indistinguishable from \(1-e^{-t/2N_e}\). Results are shown for
\(\alpha_k=1/6\). As \(\alpha_k\) decreases to 0, \(\bar{F}\) in the
mixed-ploidy population becomes completely indistinguishable from
\(\bar{F}\) in the diploid subpopulation.}\label{fig:fNe}
\end{figure}

@@

Experimental results indicate that, at least in yeast, triploids usually form
trivalents and undergo recombination, after which each trivalent is randomly
assorted in the daughter cells, some receiving one, others two copies of a
given chromosome \citep{charles2010}.
In the absence of gametic nonreduction, the probability
of obtaining euploid gametes (two diploid and two haploid gametes) from such 
a process is $(1/2)^n$, where $n$ is the number of chromosomes. If the number
of chromosomes is small this is not negligible, for instance in
\textit{Arabidopsis thaliana} we would have $(1/2)^5 \approx 0.03$, which is of
the same order as the unreduced gamete formation rate.
Unreduced (triploid) gametes may also be produced and important for the
dynamics of mixed-ploidy populations \citep{ramsey1998}.
However, they generate additional difficulty, since in order to compute the
contributed variance under inbreeding, we would need an additional identity
coefficient recording the probability that three genes are IBD at a locus.
We will hence ignore the possibility of unreduced gamete production in
triploids.
We note that, on the supposition that diploid gametes are produced by random
assortment of chromosomes in a haploid and diploid gamete, $\alpha_3 \le 1/4$.

@@

\subsection{Equilibrium trait value distribution for a large mixed-ploidy
population \label{sec:mixeq}}

In our establishment model, we assume that migrant individuals are drawn
randomly from a large (effectively infinite) non-inbred mixed-ploidy population
at HWLE and cytotype equilibrium, with mean trait value zero.
The equilibrium trait value distribution in a monocytotypic population
under such conditions is straightforwardly obtained under the infinitesimal
model as a Gaussian with variance $k \beta_k^2 V$ (see main text).
However, the equilibrium trait value distributions for the different cytotypes
in the mixed-ploidy model are more challenging to obtain.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/mixeq.pdf}
\caption{
    Equilibrium trait value distribution for diploids, triploids and
    tetraploids.
    The simulated equilibrium distribution, the small $u$ approximation and the
    monocytotypic equilibrium distribution are shown for all three cytotypes.
    The diploid monocytotypic distribution $\Normal(0,2V)$ is shown everywhere
    as a reference. Here we assumed $\beta_3 = \beta_4 = 1$ and $\alpha_2 =
    1/2, \alpha_3 = 1/4$ and $\alpha_4 = 1/6$.
\label{fig:mixeq}}
\end{figure}

Indeed, each cytotype produces haploid and diploid gametes with Gaussian
gametic values, distributed according to \cref{m-eq:gamete} in the main text.
Conditional on the parental trait values, offspring trait values will be sums
of independent Gaussians, and hence again Gaussian, distributed according to
\cref{m-eq:oneline} in the main text.
Consider now, for instance, a triploid individual, there are six different
parental cytotype combinations that may yield a triploid offspring, with
relative frequencies proportional to the terms in the expansion of the
following polynomial in $\pi$
  $$2((1-u)\pi_2 + v\pi_3)(u\pi_2 + v\pi_3 + (1-u)\pi_4)$$
(where the term in $\pi_2^2$ is proportional to the frequency of triploids
derived from two diploid parents, the term in $\pi_2\pi_3$ proportional to the
frequency of triploids derived from a diploid and a triploid parent,
\textit{etc.}).
Each of these combinations of parental cytotypes potentially yields a different
Gaussian distribution over offspring trait values.
As a result, the triploid trait distribution in the next generation will be a
\textit{mixture} of Gaussians, with mixture proportions given by the relative
frequencies in the expansion of the above polynomial.
A similar reasoning applies to the diploid and tetraploid subpopulation.
Proceeding to the next generation, the trait distributions will be mixtures
mixtures of Gaussians, and so on for further generations.

We have not identified whether this process yields a tractable equilibrium
distribution. However, for a predominantly diploid population with $u$
sufficiently small, an accurate approximation is readily obtained by assuming
that effectively all individuals in the population have diploid parents.
In this case, still assuming the mean trait value to be zero, diploid trait
values will be Gaussian with variance $2V$.
Using \cref{m-eq:oneline}, the trait value distribution of newly formed triploids
will be Gaussian with mean zero and variance
\begin{align*}
  \Var[Z_{ij}^{12}] 
    &= \Ex[\Var[Z_{ij}^{12}|Z_i,Z_j]] + \Var[\Ex[Z_{ij}^{12}|Z_i,Z_j]] \\
    &= \beta_3^2\left(\frac{V}{2} + 2\alpha_2 V\right) +
    \Var\left[\beta_3\left(\frac{Z_i}{2} + Z_j\right)\right] \\
    &= 3\beta_3^2V\left(1 + \frac{2}{3}\alpha_2\right)
\end{align*}
(where $Z_{ij}^{12}$ denotes the trait value of a random offspring from
individuals $i$ and $j$, and where the superscript indicates that $i$
contributes a haploid gamete, and $j$ a diploid gamete -- as in
\cref{m-eq:oneline}).
Under the small $u$ assumption, $Z_i$ and $Z_j$ are random draws from a
Gaussian with mean zero and variance $2V$.
The trait value distribution for tetraploids will be Gaussian with mean zero
and variance
  $$\Var[Z_{ij}^{22}] = 4\beta_4^2V\left(1 + \alpha_2\right)$$
This approximation is illustrated in \cref{fig:mixeq}, where, the fit to
simulations is shown to be very accurate.
Note that when the $\alpha_k$ parameters are set to zero (as in most of our
results), the equilibrium variance for each cytotype in the mixed-ploidy
population is just the equilibrium variance in an isolated monocytotypic
population.

\bibliographystyle{abbrvnat}
\bibliography{/home/arzwa/vimwiki/bib.bib}
\end{document}
